A novel route to the spontaneous formation of porous crystals via viscoelastic phase separation

### 2016-03-13 nature/response.tex

\begin{document}

\doublespacing

\doublespacing

\noindent
Dear Editor,
\vskip 0.3cm
we would like to thank you for your comments to our manuscript ``A novel route to the spontaneous formation of porous crystals via viscoelastic phase separation'' (NM15113574),
and for providing us with three Referee reports.
We were happy to see that all Referees appreciated our work, and at least two of them recommended it after some technical clarifications.
For this new version of the manuscript we have conducted additional experiments and analysis,
which we believe address all points raised by the Referee.

Referee 1 describes our results as ``very nice and beautiful'' but does not recommend our work
based on the claim that it is not novel. He refers in particular to a series of works by the Edinburgh group in which the kinetics of phase
transitions is classified according to the free energy landscape. These work are by themselves very important and relevant, but contrary
to the Referee claim, do not contain a description or results on the crystallization channel we describe in our work.
Both Referees 2 and 3 have appreciated this fact.

Referee 2 recommended our work, and had only minor technical questions. The most important one was regarding the phase boundaries
which were computed with free-volume theory. In the present manuscript we have strengthened our results with a Gaussian field analysis
that shows full consistency between theory and experimental observations.

Referee 3 main concern comes from her/his direct inspection of our supplemental movies, where she/he observed an anomalous abundance
of gas particles, that led her/him to suggest that Ostwald ripening was the main crystallization channel. But this observation
was misguided by the fact that the two supplemental movies are shot at very different framerates. This was explained in the movies captions,
but there was no time stamp on the movies, and we believe that this was the source of confusion, for which we apologize. While the first movies covers
the first minutes of the process, the second movie shows the evolution over more than 7 hours, giving the impression
of an abrupt increase of gas particles between the two. Also the movies display only a thin slice of the sample, inside which
the crystalline nuclei appear to be separate from each other, which led the Referee to suggest Ostwald ripening. But looking at the process in three dimensions,
it is without any doubt that crystalline nuclei are connected in a porous network. We show this very clear three-dimensional representation in a new Supplementary Figure.
Together with a novel analysis which determines the crystallization pathway directly by looking
at the particle trajectories, we believe that the Referee's issue is now fully solved.

@@

\singlespacing

{\bf
This manuscript describes experimental work on a colloid-polymer mixture that undergoes viscoelastic phase separation thereby forming a spinodal structure, which is interrupted by crystallization. The authors argue that this mechanism can be used to form nano-porous crystals, which has potential applications. However, this phenomenon has been shown in extensive previous studies, see Ref. [7-10], and is therefore not novel. Another key result of the paper is that the authors demonstrate that the crystalline branches of the spinodal network can also grow by the attachment of "gas"-like particles onto the surfaces. Even this message is not completely new as the Edinburgh group (Poon, Evans, Renth, Cates) has made extensive studies on the precise kinetic pathways of the phase transformations in colloid-polymer mixtures and explained this by means of free-energy landscapes. An extensive body of work that the authors apparently missed. Although the experimental study and analysis is very nice and beautiful, the 
referee hesitate to recommend the manuscript for publication in Nature Materials due to a lack of novelty (in terms of material and fundamental insight) of the present work.
}

\bigskip
\doublespacing

First we would like to thank the Referee for her/his careful reading of our manuscript, for her/his comments.

@@

\singlespacing

{\bf
1) The evidence for this crystal gel mechanism appears to derive from measurements at a single state point. For this sample, what happens at the transition between the two SI movies, at about 500 Brownian times? At early times, I see very few free particles, and no crystallization. A few frames into the longer movie, I see many more free particles, with exchange between the dense and dilute regions, mostly from one crystallite to the other. What, within the scenario proposed, would lead to such an abrupt increase in the abundance of free particles at ~ 500 Brownian times? It is as if the system itself changed its state point and shifted over a boundary that then allowed the crystallization to accelerate. (The increase in free particles occurs more at ~ 500 Brownian times occurs more rapidly than I think would be expected based on the kinetics of spinodal decomposition.) I think that additional observation of the phenomena of interest at different state points and/or evidence for replication at the indicated 
state points are needed to rule out idiosyncratic (time-dependent) physical chemical explanations for the abrupt increase in gas phase colloids.
}

\bigskip
\doublespacing

We think that we have not been clear enough about the explanations of the Supplementary movies. Movie 1 and Movie 2 were made with different frame rates and there was an interval of 90 s between them. This is because the dynamics is rather fast in the initial stage driven by spinodal decomposition. But, in the late stage, the dynamics slows down and thus we reduced the frame rate of 3D scan to cover the long-time period. We are afraid that the Reviewer’s impression of a sudden increase in free gas particles and a possible discontinuity between the two movies originatse from this difference in the frame rate.
To address this issue we have now included time stamps on both Supplemental Movies.

To address the concern about the reproducibility of our kinetic pathway, we decided to make additional independent experiments
by using a sample with similar composition, which is inserted into an ordinary glass capillary tube. We thus confirm the reproducibility of this type of crystallization pathway. An example of crystal-gel porour network obtained through our new capillary experiments
is found in Fig. SXXX. All details about the experiments are included the Supplementary Information. We would like to thank the Referee for stimulating
this additional investigation.

@@

{\bf
2) From the SI movie, I don't recognize the final reported structure to be a crystal gel. These final structures seem instead to be crystallites of finite size at equilibrium. The crystallites diffusive minimally because they are jammed against their neighbors. The results seem similar to Anderson and Lekkerkerker's (ref 20) explanation of Ostwald ripening. The changes in crystal size at late times, when the liquid to gas to crystal mechanism is proposed, appear that they could be equally well described as a process in which larger crystallites grow at the expense of smaller ones, through the gas phase.

}

\bigskip
\doublespacing

The impression of the reviewer that the final structures are made of crystallites of finite size may come from the fact that the percolated network structure cannot be seen by 2D sectioning. The Movies show only a thin slice of the system which corresponds to the particle that are in focus.
To show the connectivity of the network clearly, we have also made Supplementary Figure SXXX, highlighting the crystal network.
We also include movies of a thin slice with a finite thickness. \textcolor{red}{Did we do this?} 
We hope that this will clarify the situation and convince the reviewer of the fact that our sample indeed forms a porous network of crystals.

@@

\vspace{1em}

\singlespacing

{\bf
4) In panel 4b, about ~ 25\% of the particles are identified as in the surface state; however, transition probabilities are not recorded for this state in Figure 4c. From the SI movies, it seems that most of the kinetics are between surface and gas states at long times (as in Ostwald ripening). I did not find explanation in the paper about how these transition probabilities are defined and measured, and how the surface particles are excluded from the analysis. Also, do the kinetics measured satisfy detailed balance at long times? These data should include crystal-gas and crystal-liquid transition probabilities so that the author's mechanism of liquid to gas to crystal can be evaluated against the alternative that the sublimination/desublimination transitions are from crystal to gas to crystal.
}

In the previous version of the manuscript, Fig. 4c included surface particles together with crystalline particles just to improve
the readability of the graph, but we agree with the Referee that by doing so we didn't convincingly show that the Bergeron channel (liquid-to-gas)
is more prominent than the Ostwald ripening channel (crystal-to-gas). 
To include the full analysis, as requested by the Referee,
we have added a section to Supplementary Information, with a new figure (Fig. S3) containing four panels that detail the transition
probabilities between all states. These transition probabilities are measures by reconstructing the particles' trajectories, and
computing probability of transition between different states at each time.
In particular we note that, while being an active channel, the Ostwald ripening pathway is limited by
the small conversion probability of surface particles into the gas phase.

### 2016-03-13 crystal-gel/nature/crystal_gel2.tex
Small addition to text.

\begin{figure*}[!t]
 \centering
 \includegraphics{phase_separation.pdf}
 \caption{{\bf Different routes of viscoelastic phase separation.} 
{\bf a,} Phase diagram. Empty black triangles are experimental point showing a fluid behaviour. The other symbols are the experimental points showing gel behaviour, consistently used in all panels. The dotted lines are the gas-liquid tie lines determined experimentally, with arrow heads indicating measured gas and liquid compositions. Gray areas are theoretical equilibrium fluid and crystal regions. The solid and dashed lines are the metastable gas-liquid binodal and spinodal respectively.
{\bf b,} Radius of gyration ($R_g$) as a function of cluster size ($s$) for all state points at percolation. The
 gray line represents the fractal dimension of the random gelation universality class.
{\bf c,} Structure factors for the late stages of the gelation process. The gray and black lines represents respectively random gelation and compact fractal dimensions.
{\bf d,} Local volume fraction around a colloidal particle, as a function of the number of nearest neighbours of the particle for the state points in the very late stage. Horizontal lines stress the value at 12 neighbours.
 }
 \label{fig:phase_separation}
\end{figure*}

>>>>

\begin{figure*}[!t]
 \centering
 \includegraphics{phase_separation.pdf}
 \caption{{\bf Different routes of viscoelastic phase separation.} 
{\bf a,} Phase diagram. Empty black triangles are experimental point showing a fluid behaviour. The other symbols are the experimental points showing gel behaviour, consistently used in all panels. The dotted lines are the gas-liquid tie lines determined experimentally, with arrow heads indicating measured gas and liquid compositions (see Supplementary Information on the procedure to extract the compositions from confocal images).  Gray areas are theoretical equilibrium fluid and crystal regions. The solid and dashed lines are the metastable gas-liquid binodal and spinodal respectively.
{\bf b,} Radius of gyration ($R_g$) as a function of cluster size ($s$) for all state points at percolation. The
 gray line represents the fractal dimension of the random gelation universality class.
{\bf c,} Structure factors for the late stages of the gelation process. The gray and black lines represents respectively random gelation and compact fractal dimensions.
{\bf d,} Local volume fraction around a colloidal particle, as a function of the number of nearest neighbours of the particle for the state points in the very late stage. Horizontal lines stress the value at 12 neighbours.
 }
 \label{fig:phase_separation}
\end{figure*}
